---
title: "In-Class_Exercise 07"
author: "Priya"
---

##Getting started

```{r}
pacman::p_load(tidyverse, tsibble, feasts, fable, seasonal)
```

```{r}
ts_data <- read_csv("data/visitor_arrivals_by_air.csv")
```

```{r}
ts_data$'Month-Year' <- dmy(ts_data$'Month-Year')
```

### Conventional base 'ts' object versus 'tibble' object

ts object

```{r}
ts_data_ts <- ts(ts_data)
head(ts_data_ts)
```

### **Converting `tibble` object to `tsibble` object**

```{r}
ts_tsibble <- ts_data %>%
  mutate(Month = yearmonth(`Month-Year`)) %>%
  as_tsibble(index = `Month`)
```

## **Visualising Time-series Data**

```{r}
ts_longer <- ts_data %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")

```

### **Visualising single time-series: ggplot2 methods**

```{r}
ts_longer %>%
  filter(Country == "Vietnam") %>%
  ggplot(aes(x = `Month-Year`, 
             y = Arrivals))+
  geom_line(size = 0.5)
```

### **Plotting multiple time-series data with ggplot2 methods**

```{r}
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals,
           color = Country))+
  geom_line(size = 0.5) +
  theme(legend.position = "bottom", 
        legend.box.spacing = unit(0.5, "cm"))
```

```{r}
#| echo: false
#| fig-height: 12
ggplot(data = ts_longer, 
       aes(x = `Month-Year`, 
           y = Arrivals))+
  geom_line(size = 0.5) +
  facet_wrap(~ Country,
             ncol = 4,
             scales = "free_y") +
  theme_bw()
```

### **Visual Analysis of Seasonality with Seasonal Plot**

```{r}
tsibble_longer <- ts_tsibble %>%
  pivot_longer(cols = c(2:34),
               names_to = "Country",
               values_to = "Arrivals")
```

```{r}
#| fig-height: 6
tsibble_longer %>%
  filter(Country == "Italy" |
         Country == "Vietnam" |
         Country == "United Kingdom" |
         Country == "Germany") %>% 
  gg_season(Arrivals)
```

### **Visual Analysis of Seasonality with Cycle Plot**

```{r}
tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  autoplot(Arrivals) + 
  facet_grid(Country ~ ., scales = "free_y")





```

```{r}
#| fig-height: 6
tsibble_longer %>%
  filter(Country == "Vietnam" |
         Country == "Italy") %>% 
  gg_subseries(Arrivals)
```

## **Visual Forecasting**

```{r}
vietnam_ts <- tsibble_longer %>%
  filter(Country == "Vietnam") %>% 
  mutate(Type = if_else(
    `Month-Year` >= "2019-01-01", 
    "Hold-out", "Training"))
```

```{r}
vietnam_train <- vietnam_ts %>%
  filter(`Month-Year` < "2019-01-01")
```

```{r}
vietnam_train %>%
  model(stl = STL(Arrivals)) %>%
  components() %>%
  autoplot()
```

### 

```{}
```
